Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-01-21 grove-paper.tex
fleshing out decomposition algorithm

The language we are modeling is shown in \autoref{fig:Base syntax}, standard for
a simply-typed lambda calculus except that all sorts have empty holes
(represented by $\hole$) and conflict holes (of the form $\{q | \ldots | q\}$).
In particular, the argument of a lambda can be either a bound variable or a
hole. Additionally, all non-hole forms carry a superscripted identifier $u \in
\U$.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \hfil \rightarrow & \wp(\P)                                        \\
      \hline
      \arity(\Root)={}        & \left\{ \Root \right\}                         \\
      \arity(\PatVar)={}      & \left\{ \right\}                               \\
      \arity(\ExpVar)={}      & \left\{ \right\}                               \\
      \arity(\ExpLam)={}      & \left\{ \mathtt{Param}, \mathtt{Type}, \mathtt{Body} \right\} \\
      \arity(\ExpApp)={}      & \left\{ \mathtt{Fun}, \mathtt{Arg} \right\}              \\
      \arity(\ExpPlus)={}     & \left\{ \mathtt{L}, \mathtt{R} \right\}         \\
      \arity(\ExpTimes)={}    & \left\{ \mathtt{L}, \mathtt{R} \right\}         \\
      \arity(\ExpNum)={}      & \left\{ \right\}                               \\
      \arity(\TypArrow)={}    & \left\{ \mathtt{Arg}, \mathtt{Result} \right\}       \\
      \arity(\TypNum)={}      & \left\{ \right\}                               \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

A vertex $v = (u, k)$ identifies a distinct instance of some constructor $k \in
\K$, corresponding to a form defined in \autoref{fig:Syntax as a grammar} that
is neither an operator nor a hole. Each contructor has an arity, which we define
as a set of positions. A position $p \in \P$ identifies a location, adjacent to
some source vertex, from which edges to target vertices may originate. The
$\arity$ function in \autoref{fig:Constructors, Indexes and Arity} gives the
complete mapping from constructors to potential edge positions. Every graph
contains a distinguished root vertex, root constructor, and root position,
respectively named $v_0$, $k_0$, and $p_0$.

% TODO: should graph "recomposition" have an algorithm, too?
A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and constructor $k$ for the outermost
expression, and creating edge $\e = (u, v_0, p_0, (u_k, k))$. Then, for each
sub-term at some position $p \in \arity(k)$, select unique id $u'$ and
constructor $k'$ and create edge $e' = (u', v, p, v')$ with $v = (u_k, k)$, $v'
= (u', k')$. Continuing in this manner, treat each sub-term, starting with the
ones at positions in $\arity(k')$, as if it were the outermost and create edges
to the vertices representing its immediate sub-terms until only empty holes
remain. Note that empty holes are not modeled explicitly in the graph. Wherever
an empty hole appears in an expression, the corresponding graph position has no
edges originating from it.

@@

% TODO: clean up and move rooted expression grammar

@@

(illustrative running example: 1 of each: mp, d, r, sc)

@@

Tree-like structures are often easier to work with than arbitrary graphs
directly, for example to present the graph to users, or for type checking, so we
decompose it into a grove.

A grove is a 4-tuple of the following form:
\[
  (e, MP, NP, W)
\]
The first component is an expression, governed by the grammar in
\autoref{fig:Base syntax}, determined by starting at the root of the graph and
including every connected vertex that has a single parent. In positions where
there are no children, we leave an empty hole, $\hole$. In positions where there
are multiple children, we leave a conflict hole of the form $\{e_1 | \ldots |
e_n\}$. Whenever we encounter a vertex with multiple parents, we refer to it
indirectly by reference $\#u$. We collect the expressions corresponding to
vertices with multiple parents in the set $MP$.

These first two components do not account for vertices with no parents. We
include the expressions corresponding to these vertices in the set $NP$.

We've accounted for vertices with multiple parents, with no parents, and with
one parent connected to the root. This leaves only vertices with one parent that
are not connected to the root. These vertices form structures we call wreaths.

Wreaths:

Three components are  <describe each of the components:
The first component is ... trees. The fourth component is ... wreaths>.

A forest is a collection of trees, in fact it is a 4-tuple consisting of
a rooted tree $R$ and 3 sets of trees: $MP, D, W$. Before discussing what these
sets are, let's define what a tree is.

here's what a tree is. why do we care what a tree is? because we can map them to
the expressions in figure 11. Explain what a Ref is wrt forest roots. (Maybe
mention that only well-sorted trees can be mapped, which we discuss below.)

There's this weird stuff left over; we call them wreaths and put them in the 4th
tuple component.

here's what a wreath is (= a connected graph where each vertex has exactly one
parent)

(show what a wreath looks like)

here's what a grove is

here are some properties that show how to get one from the other, and to/from
graph

>>>>

The language we are modeling is shown in \autoref{fig:Base syntax}, standard for
a simply-typed lambda calculus except that all sorts have empty holes
(represented by $\hole$) and conflict holes (of the form $\{q | \ldots | q\}$).
In particular, the argument of a lambda can be either a bound variable or a
hole. Additionally, all non-hole forms have a superscripted identifier $u \in
\U$.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{ll}
      \arity : \K \hfil \rightarrow & \wp(\P \times \{Exp, Pat, Typ\}) \\
      \hline
      \arity(\Root)={} & \left\{ \Root \right\} \\
      \arity(\PatVar)={} & \left\{ \right\} \\
      \arity(\ExpVar)={} & \left\{ \right\} \\
      \arity(\ExpLam)={} & \left\{ (\mathtt{Param}, Pat), (\mathtt{Type}, Typ), (\mathtt{Body}, Exp) \right\} \\
      \arity(\ExpApp)={} & \left\{ (\mathtt{Fun}, Exp), (\mathtt{Arg}, Exp) \right\} \\
      \arity(\ExpPlus)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpTimes)={} & \left\{ (\mathtt{L}, Exp), (\mathtt{R}, Exp) \right\} \\
      \arity(\ExpNum)={} & \left\{ \right\} \\
      \arity(\TypArrow)={} & \left\{ (\mathtt{Arg}, Typ), (\mathtt{Result}, Typ) \right\} \\
      \arity(\TypNum)={} & \left\{ \right\} \\
    \end{array}
  \]
  \caption{Constructors, Indexes and Arity}
  \label{fig:Constructors, Indexes and Arity}
\end{figure}

@@

A vertex $v = (u, k)$ identifies a distinct instance of some constructor $k \in
\K$, corresponding to a non-hole form defined in \autoref{fig:Base syntax}. Each
contructor has an arity, which we define as a set of positions. A position $p
\in \P$ identifies a location at some originating vertex from which edges to
destination vertices may originate. The $\arity$ function in
\autoref{fig:Constructors, Indexes and Arity} gives the complete mapping from
constructors to potential edge positions with their corresponding sorts. Every
graph contains a distinguished root vertex, root constructor, and root position,
respectively named $v_0$, $k_0$, and $p_0$.

% TODO: should graph "recomposition" have an algorithm, too?
A graph can be derived from an expression in the calculus by starting with an
empty graph, selecting unique ids $u, u_k$ and constructor $k$ for the outermost
expression, and creating edge $\e = (u, v_0, p_0, (u_k, k))$. Then, for each
sub-term at some position $p \in \arity(k)$, select unique id $u'$ and
constructor $k'$ and create edge $e' = (u', v, p, v')$ with $v = (u_k, k)$, $v'
= (u', k')$. Continuing in this manner, treat each sub-term as if it were the
outermost, starting with the ones at positions in $\arity(k')$, and create edges
to the vertices representing their immediate sub-terms until only empty holes
remain. Note that empty holes are not modeled explicitly in the graph. Wherever
an empty hole appears in an expression, the corresponding graph position has no
edges originating from it.

@@

@@

@@

Trees are often easier to work with than arbitrary graphs directly, for example
to present the graph to users or for type checking. A graph is not necessarily a
tree, but we can decompose any graph into a collection of tree-like structure we
call a grove.

A grove is a 4-tuple of the following form:
\[
  (e, MP, NP, W)
\]
The first component $e$ is an expression as governed by the grammar in
\autoref{fig:Base syntax}. We determine $e$ by starting at the root of the graph
and including every connected vertex that has a single parent. In positions
where there are no children, we leave an empty hole $\hole$. In positions where
there are multiple children, we leave a conflict hole of the form $\{e_1 |
\ldots | e_n\}$. Whenever we encounter a vertex with multiple parents, we leave
an indirect reference $\#u$, where $u$ is the unique identifier of the
referenced vertex, and add its corresponding expression to the second component
$MP$.

These first two components do not account for vertices with no parents. We
include the expressions corresponding to these vertices in the third component
$NP$.

We've accounted for single-parented vertices connected to the root,
single-parented vertices connected to multi-parented vertices, and vertices with
no parents. This leaves only single-parented vertices that are not connected to
the root or any multi-parented vertices. These vertices form cyclic structures
we call wreaths.

A wreath is a cycle, potentially with acyclic branches. Wreaths are connected
graphs where each vertex has exactly one incoming edge. The fourth component $W$
is the set of expressions determined by extracting wreaths from any vertices not
in the other three components.

% expression = tree

% (show what a wreath looks like)

% here are some properties that show how to get one from the other, and to/from
% graph
